---
title: "Fisher's Exact Test"
author: "Igor Starinskij"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
---

```
#06_fisher.exact.rmd

#Purpose:
#    Run a fisher exact test with Benjamini-Hochberg adjustment.

#Input:
#    Excel workbook containing a contingency table.  

#Output:
#   1. .html document with analysis flowthrough and outputs
#	2. .csv with p values pre and post multiple comparisons correction

#Thesis:
#    Chapter: ERVEV biosurveillance survey

#Author:
#    Igor Starinskij

#AI assistance:
#   Generative artificial intelligence tools, primarily GitHub Copilot 0.63.0 (GPT-5.6 Luna), were used during the development of software used in this thesis.
#	AI was used to assist with R code generation, debugging, code refinement and explanation of programming concepts.
#	Generated code was reviewed, modified, tested and validated by the author.
#	The analytical methodology, statistical decisions, interpretation of results and final responsibility for the analyses remain with the author.
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Import the workbook

Select an Excel workbook containing a sheet named `fisher`. The sheet must contain
numeric, non-negative counts arranged as a contingency table. A first non-numeric
column may be used for row labels.

```{r import}
if (!requireNamespace("readxl", quietly = TRUE)) {
	stop("This report requires the 'readxl' package. Install it with install.packages('readxl').")
}

file_path <- file.choose()
available_sheets <- readxl::excel_sheets(file_path)

if (!"fisher" %in% available_sheets) {
	stop("The selected workbook does not contain a sheet named 'fisher'.")
}

df <- readxl::read_excel(file_path, sheet = "fisher", .name_repair = "unique")

if (nrow(df) == 0L || ncol(df) == 0L) {
	stop("The 'fisher' sheet is empty.")
}

knitr::kable(df, caption = "Imported data")
```

## Validate the contingency table

The first column is treated as row labels when it is not numeric. All remaining
columns must then be numeric counts. If the first column is numeric, every column
is treated as part of the table.

```{r validate}
is_numeric_column <- vapply(df, is.numeric, logical(1))
has_label_column <- !is_numeric_column[[1L]]

if (has_label_column) {
	row_labels <- as.character(df[[1L]])
	count_data <- df[-1L]
} else {
	row_labels <- rownames(df)
	count_data <- df
}

if (nrow(count_data) < 2L || ncol(count_data) < 2L) {
	stop("The 'fisher' sheet must contain at least two rows and two columns of counts.")
}

if (!all(vapply(count_data, is.numeric, logical(1)))) {
	stop("All contingency-table cells must be numeric. Check the headers and count columns.")
}

observed <- as.matrix(count_data)
storage.mode(observed) <- "numeric"

if (anyNA(observed) || any(!is.finite(observed))) {
	stop("The contingency table contains missing or non-finite values.")
}

if (any(observed < 0) || any(observed != floor(observed))) {
	stop("Contingency-table cells must be non-negative whole-number counts.")
}

if (any(rowSums(observed) == 0) || any(colSums(observed) == 0)) {
	stop("Rows and columns must each contain at least one observed count.")
}

rownames(observed) <- row_labels
colnames(observed) <- names(count_data)

cat("Workbook:", basename(file_path), "\n")
cat("Sheet: fisher\n\n")
cat("Observed contingency table:\n")
print(observed)
```

## Expected counts and statistical test

```{r analysis}
chi_square <- suppressWarnings(stats::chisq.test(observed, correct = FALSE))
expected <- chi_square$expected

cat("Expected counts under independence:\n")
print(round(expected, 4))

cat("\nMinimum expected count:", round(min(expected), 4), "\n")
cat("Chi-square test p-value:", format.pval(chi_square$p.value), "\n")

if (any(expected < 5)) {
	    set.seed(123)
    
        fisher_result <- tryCatch(
         stats::fisher.test(observed, workspace = 2e8),
        error = function(error) {
            if (!grepl("FEXACT error 7", conditionMessage(error), fixed = TRUE)) {
                stop(error)
            }
    
            stats::fisher.test(
            observed,
            simulate.p.value = TRUE,
            B = 100000
        )
      }
    )

	cat("\nAt least one expected count is less than 5.\n")
	cat("Fisher's exact test p-value:", format.pval(fisher_result$p.value), "\n")
    cat("Test method:", fisher_result$method, "\n")

	if (fisher_result$p.value < 0.05) {
		row_test <- function(row_index) {
			row_vs_rest <- rbind(
				observed[row_index, , drop = FALSE],
				colSums(observed[-row_index, , drop = FALSE])
			)
			rownames(row_vs_rest) <- c(rownames(observed)[row_index], "All other rows")

			tryCatch(
				stats::fisher.test(row_vs_rest, workspace = 2e8),
				error = function(error) {
					if (!grepl("FEXACT error 7", conditionMessage(error), fixed = TRUE)) {
						stop(error)
					}

					stats::fisher.test(
						row_vs_rest,
						simulate.p.value = TRUE,
						B = 100000
					)
				}
			)
		}

		row_tests <- lapply(seq_len(nrow(observed)), row_test)
		row_results <- data.frame(
			row = rownames(observed),
			p_value = vapply(row_tests, function(test) test$p.value, numeric(1)),
			method = vapply(row_tests, function(test) test$method, character(1)),
			stringsAsFactors = FALSE
		)
		row_results$adjusted_p_value_BH <- p.adjust(row_results$p_value, method = "BH")
		row_results$significant_after_BH <- row_results$adjusted_p_value_BH < 0.05

		row_results_path <- file.path(
			dirname(file_path),
			paste0(tools::file_path_sans_ext(basename(file_path)), "_fisher_row_tests.csv")
		)
		utils::write.csv(row_results, row_results_path, row.names = FALSE)
		cat("\nGlobal Fisher/Monte Carlo p-value is below 0.05.\n")
		cat("Row-vs-rest results saved to:", row_results_path, "\n")
	} else {
		cat("\nGlobal Fisher/Monte Carlo p-value is not below 0.05; row-vs-rest tests were not performed.\n")
	}

	fisher_result
} else {
	cat("\nAll expected counts are at least 5; Fisher's exact test was not performed.\n")
}
```

## Reported results

```{r results, results='asis'}
cat("- Input file: `", basename(file_path), "`\n", sep = "")
cat("- Observed table dimensions: ", nrow(observed), " x ", ncol(observed), "\n", sep = "")
cat("- Minimum expected count: ", round(min(expected), 4), "\n", sep = "")
cat("- Chi-square p-value: ", format.pval(chi_square$p.value), "\n", sep = "")

if (exists("fisher_result")) {
	cat("- Fisher's exact test p-value: ", format.pval(fisher_result$p.value), "\n", sep = "")
	if (exists("row_results")) {
		cat("- Row-vs-rest results saved to: `", basename(row_results_path), "`\n", sep = "")
		knitr::kable(row_results, digits = 6, caption = "Row-vs-rest Fisher tests with Benjamini-Hochberg adjustment")
	} else {
		cat("- Row-vs-rest tests: not performed because the global p-value is not below 0.05.\n")
	}
} else {
	cat("- Fisher's exact test: not required because all expected counts are at least 5.\n")
}
```
